---
title: "SAEM_IRT Tutorial"
output: html_notebook
---


# Probit Link MIRT: A Gibbs Sampler Applying SAEM
Welcome to the tutorial on using the SAEM_IRT code to fit your IRT exploratory models.

The first thing we want to do is to appropriately set up your environment for running our code. The following code will appropriately set the working directory and demonstrate the R Notebook features if you have not yet been introduced to this markdown format.
```{r}
knitr::opts_chunk$set(echo=FALSE,warning = FALSE)
source("Hello.R", chdir = T)
Hello()
```

## R Environment and Packages
Before we get started, we'll have to make sure you have some packages installed. We used a C++ implementation for some functions so Rcpp is key. Also, for our parallelization pleasure, we'll include some logic to decide which packages are appropriate in your environment.
```{r,message=FALSE}
### Packages called ###
PACK = c(ifelse(get_os()=="windows","snow","parallel"),"foreach",
         ifelse(get_os()=="windows","doSNOW","doParallel"),
         "rlecuyer","GPArotation","mvnfast","psych","modeest","MASS",
         "Rcpp","RcppArmadillo","devtools","compiler","doRNG","abind",
         "truncdist","truncnorm","modeest","combinat","fastGHQuad","bdsmatrix")

### Install packages not already installed ###
for (i in 1:length(PACK)){
  if(PACK[i] %in% rownames(installed.packages()) == FALSE) {install.packages(PACK[i],dependencies = TRUE)}
}
# NOTE: You will not see output when message=FALSE. You will still see warnings, etc.
```


### Assumptions
In running the SAEM algorithm, you'll want to either load your own data or generate data. It is a good idea to generate some data to get a feel for the inputs and outputs, and that's what we'll do here. Several data sets are included to help facilitate your work with this code as well.

Please NOTE the following assumptions here: 

* This suite of functionality assumes orthogonal latent factors in its factor analysis.
* Rotations on output are made based on assumptions about the structure of multidimensionality in items.
* The model uses a probit link. Later in this tutorial, we will transform back to a 2PL or 3PL for pedagogical purposes.
  
## Generate Data
A very simple approach is to generate a small dichotomous test. Let's say 10 items, 500 examinees, and let's keep it simple with a single slope and intercept per item. The latent factor will be a normal distribution, there will be no guessing, and the model of the item characteristic curve will be a probit. There are many other parameters required for properly sampling
```{r}
J = 10; N <- 500; Q <- 1; K <-NA
structure=list(icc="ogive",          # Item Char Curve; "ogive" or "logistic" 
               Adist="beta",         # prior distribution of A's/loadings
               Aparams=c(0.2,1.7),   # parameters of A's/loadings' prior distribution
               Adim=Q,               # 1 (univariate) or 2, 3, etc. multiple dimensions for multivariate
               bdist="norm",         # distribution of B/intercept
               bparams=c(0,1),       # parameters of bdist
               guess=FALSE,          # guessing ? TRUE/FALSE
               ncat=K,               # Ordinal Polythomous? Number of categories
               taudist="norm",       # sample distribution for categorical intercepts
               tauparams=c(0,1),     # parameters for taudist
               cdist="unif",         # guessing parameter distribution for 3PNO or 3PL
               cparams=c(0.05,0.3),  # bounds
               tmu=rep(0,Q),         # Theta Prior... e.g. 0, or multivariate c(0,0) ... can be multidimensional
               tsigma=if(Q==1) 1 else diag(Q), # Latent factor orthogonal covariance
               simfile=paste0("Poly_J",J,"_N",N,"_Q",Q,ifelse(!is.na(K) & K>2,paste0("_K",K),""),".rda"))  # Name of saved file of generated data.

# This will show you the parameters you're using for sampling the response pattern.
VisualizeParams(structure)
```

Now let's generate some response pattern data!
```{r}
Gen.Data<-GenerateTestData(j=J,n=N,structure=structure)
cat("\nParameters (A and b)\n")
print(head(Gen.Data$XI,n=10))
cat("\nParameters (Theta)\n")
print(head(Gen.Data$THETA))
OgiveICC(xi = Gen.Data$XI,guess=structure$guess,j=1)
cat("\nResponse Data\n")
print(head(Gen.Data$RP))
```

## Fit the Data
Now that we have response data, we want to fit it using our Gibbs Sampler.
```{r}

```

